---
title: "Custom tabulations from the Borealis Dataverse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom tabulations from the Borealis Dataverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = nzchar(Sys.getenv("COMPILE_VIG")))
```

The official StatCan census index (see `vignette("getting-started")`) is not the
only place Beyond 20/20 tables live. The
[**Borealis Dataverse**](https://borealisdata.ca) hosts a large collection of
**custom StatCan tabulations** — cross-tabulations produced on order and
deposited by researchers, sitting outside the standard product line.
Unfortunately many are simply re-posts of standard-release StatCan data, which
clogs up the Borealis repository and makes discovering the truly new tables
cumbersome, but ~3,000 are custom tables found nowhere else. `canivt` can browse
that catalogue and download from it exactly like the StatCan one.

This vignette browses the Borealis catalogue, downloads one custom tabulation,
and draws a graph from it.

```{r setup, message = FALSE}
library(canivt)
```

## An API key

The Borealis **search** (building the catalogue) needs a free API key. Create one
from your Borealis account
([API token tab](https://borealisdata.ca/dataverseuser.xhtml?selectTab=apiTokenTab))
and put it in `.Renviron` (or set it for the session):

```{r key, eval = FALSE}
Sys.setenv(BOREALIS_DATAVERSE_KEY = "xxxxxxxx-xxxx-xxxx-xxxx-xxxxxxxxxxxx")
```

The key is needed to *build* the catalogue and to *download* a file.

## 1. Browse the Borealis catalogue

`borealis_ivt_catalogue()` searches the Dataverse for every deposited `.ivt` and
returns one row per file. The search is slow (~90 s, thousands of files), so it is
cached after the first call; pass `refresh = TRUE` to rebuild it.

```{r browse, eval = FALSE}
catl <- borealis_ivt_catalogue()
dim(catl)
#> [1] 3300   11
names(catl)
#>  [1] "key"           "file"          "file_stem"     "file_id"
#>  [5] "dataset_doi"   "dataset_name"  "download_url"  "size_in_bytes"
#>  [9] "md5"           "published_at"  "file_type"
```

Each file is addressed by a canonical `key` — `<doi-token>/<file-stem>` — because
the same filename is reused across datasets (so it folds in the deposit's DOI to
stay unique). The numeric `file_id` is always unique on its own.

Filter it like any tibble. Here we look for a 2006 British Columbia custom
tabulation of housing tenure by selected characteristics (order `cro0172986`):

```{r find, eval = FALSE}
library(dplyr)
hits <- catl |>
  filter(grepl("cro0172986", file)) |>
  select(key, file, file_id, size_in_bytes)
hits
#> # A tibble: 2 × 4
#>   key                          file                                  file_id size_in_bytes
#>   <chr>                        <chr>                                 <chr>           <dbl>
#> 1 …/cro0172986_ct.7-2006-popu… cro0172986_ct.7-2006-population.ivt   …              995468
#> 2 …/cro0172986_ct.8-2006-priv… cro0172986_ct.8-2006-private_hous…    …             1358173
```

(The `key` DOI token and numeric `file_id` are the deposit's real identifiers —
shown abbreviated here; your `borealis_ivt_catalogue()` will fill them in.)

## 2. Download and decode one

Pass a one-row catalogue slice to `borealis_ivt_download()`. It fetches the file
from the Dataverse access API into the ivt cache and returns the local `.ivt`
path, which `read_ivt()` decodes like any other table:

```{r download, eval = FALSE}
path <- borealis_ivt_download(hits[1, ])   # or a bare file_id
tab  <- read_ivt(path)
tab
#> ── IVT table NA ──
#> NA
#> 102839 cells | 581 geographies | 3 dimensions | 2 footnotes
#> geography labelled by name + uid
```

(This particular custom order carries no inline product id or title — hence the
`NA` header — but the data and the dimension structure decode fine, as we'll see.)

`get_statcan_ivt()` is the one-stop alternative: give it a Borealis `key` or
`file_id` and it downloads, decodes, caches the Parquet and returns an Arrow
dataset — it resolves an id against the StatCan catalogue first and falls back to
Borealis, so the same call works for either source.

```{r get, eval = FALSE}
ds <- get_statcan_ivt("cro0172986_ct.7-2006-population")   # by file stem / key
```

The table is `Geography (581 BC census divisions & subdivisions) × Tenure
(Owner / Renter / Band housing) × Characteristics (79)`, with the population in
private households in `value`.

## 3. Query and graph

We pull the owner- vs renter-occupied **population** (the "Total - Sex of the
population" characteristic) for a set of BC cities:

```{r query, eval = FALSE}
cities <- c("Vancouver, CY", "Surrey, CY", "Burnaby, CY", "Richmond, CY",
            "Victoria, CY", "Kelowna, CY", "Abbotsford, CY", "Nanaimo, CY",
            "Kamloops, CY", "Prince George, CY")

res <- ivt_tidy(tab) |>
  filter(geo_name %in% cities,
         characteristics == "Total - Sex of the population",
         tenure %in% c("Owner", "Renter")) |>
  select(geo_name, tenure, value)
```

For a self-contained example, those real decoded 2006 counts are inline here:

```{r data, echo = FALSE}
res <- data.frame(
  geo_name = rep(c("Abbotsford, CY", "Burnaby, CY", "Kamloops, CY", "Kelowna, CY",
                   "Nanaimo, CY", "Prince George, CY", "Richmond, CY", "Surrey, CY",
                   "Vancouver, CY", "Victoria, CY"), each = 2),
  tenure = rep(c("Owner", "Renter"), times = 10),
  value = c(94705, 27590, 132865, 67330, 62145, 17020, 78760, 25805, 57180,
            19615, 52465, 17725, 136080, 37240, 307330, 84570, 324940, 244200,
            33500, 40855)
)
```

A quick tidy — the owner share of each city's household population — and an
ordered bar chart. In 2006 Surrey and the Fraser Valley are strongly
owner-occupied, while central Vancouver and Victoria are the most renter-heavy:

```{r plot, fig.width = 7, fig.height = 4.2, fig.alt = "Ordered horizontal bar chart of owner-occupancy share of the household population in ten BC cities in 2006, with a 50% reference line. Surrey, Abbotsford and Richmond are around three-quarters owner-occupied; Vancouver is just above half and Victoria is below, the only city where renters outnumber owners."}
library(ggplot2)
library(dplyr)

res |>
  mutate(city=sub(", CY$", "", geo_name)) |>
  mutate(share=value/sum(value),.by=city) |>
  filter(tenure=="Owner") |>
  ggplot(aes(share, reorder(city,share))) +
  geom_col(fill = "#2c7fb8", width = 0.72) +
  geom_vline(xintercept = 0.5, linetype = 2, colour = "grey40") +
  scale_x_continuous(labels = function(x) paste0(x * 100, "%"),
                     limits = c(0, 1), expand = expansion(mult = c(0, 0.02))) +
  labs(y = NULL, x = "Share of owner households",
       title = "Home ownership by city, B.C. 2006",
       caption = "Data: Custom StatCan tabulation on the Borealis Dataverse (cro0172986)")
```

## Recap

```r
library(canivt); library(dplyr); library(ggplot2)

catl <- borealis_ivt_catalogue()                 # 1. browse (needs an API key)
hits <- catl |> filter(grepl("cro0172986", file))

path <- borealis_ivt_download(hits[1, ])         # 2. download one custom table
tab  <- read_ivt(path)                           #    decode it like any .ivt

res  <- ivt_tidy(tab) |> filter(...)             # 3. query + graph
ggplot(res, ...) + geom_col()
```

Borealis and the StatCan catalogue are complementary sources for the same reader:
whichever a table comes from, `read_ivt()` decodes it through one path. See
`?borealis_ivt_catalogue` and `?get_statcan_ivt` for the full API, and
`vignette("getting-started")` for the StatCan-catalogue workflow.
