---
title: "Getting started: reading an older census table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: reading an older census table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = nzchar(Sys.getenv("COMPILE_VIG")))
```

This vignette walks through a typical use case: **look up a table in
the Statistics Canada catalogue, download and decode it, and work with the data**.

This example uses an **old** table that is not available in any alternative
format: the **1981 Census Profile for Canada, Provinces, Territories, Census
Divisions and Census Subdivisions** (`97-570-X1981004`). Modern census tables
come with a tidy CSV and a Web Data Service API, and more recent vintages are
also available in SDMX format, but the pre-2001 vintages are published **only**
as *Beyond 20/20* `.ivt` files — there is no CSV, no API, no easy-to-parse
alternative. `canivt` decodes them straight from the bytes, so a 1981 profile
becomes an ordinary data frame.

```{r setup, message = FALSE}
library(canivt)
```

## (Optional) set a cache

`canivt` can cache downloaded `.ivt` files and the decoded Parquet so the second
run is instant. Point it at a directory (e.g. in `.Renviron` via
`CANIVT_IVT_CACHE` / `CANIVT_DATA_CACHE`, or per session):

```{r cache, eval = FALSE}
options(
  canivt.ivt_cache  = "~/canivt-cache/ivt",   # raw .ivt downloads
  canivt.data_cache = "~/canivt-cache/data"   # decoded Parquet + catalogue
)
```

Without a cache everything still works; it just re-downloads each session. The
IVT cache is not needed if the Parquet files are the primary target — omitting
it reduces the storage footprint without impeding work with the cached Parquet
files.

## 1. Look up the table in the catalogue

`statcan_ivt_catalogue()` scrapes the StatCan census datasets index into one
tibble (one row per published table version), cached as Parquet after the first
call. Each row carries the `catalogue` number, `census_year`, `title`, `topic`
and the download URLs. The package ships with a cached version for convenience,
but this can be replaced with an up-to-date scrape by passing the
`refresh = TRUE` option.

```{r catalogue}
catl <- statcan_ivt_catalogue()

# Find the 1981 census-division / subdivision profile
library(dplyr)
catl |>
  filter(census_year == 1981, grepl("Profile.*Census Divisions", title)) |>
  select(catalogue, census_year, title)
```

If you already know the product id you can skip this step entirely and go
straight to step 2.

## 2. Download and decode with `get_statcan_ivt()`

`get_statcan_ivt()` resolves the id against the catalogue, downloads the `.ivt`,
decodes it, caches the result as Parquet, and hands back an
[`arrow`](https://arrow.apache.org/docs/r/) dataset connection. The second call
skips the download and decode.

```{r get}
ds <- get_statcan_ivt("97-570-X1981004")
ds
```

The table is a classic profile: one row per **geography × profile
characteristic**, with the count in `value`. The `profile` column holds the ~80
characteristics (population, age groups, mother tongue, tenure, …) and `geo_name`
the bilingual place name.

```{r explore, eval = FALSE}
ds |> distinct(profile) |> collect() |> head(6)
```

## 3. Query the data you want

Because `ds` is an Arrow dataset you can filter and select with `dplyr` verbs and
only `collect()` the rows you need. Here we pull the **mother-tongue** counts
(English / French / Other) for the ten provinces:

```{r query}
provinces <- c(
  "Newfoundland - Terre-Neuve", "Prince Edward Island - Île-du-Prince-Édouard",
  "Nova Scotia - Nouvelle-Écosse", "New Brunswick - Nouveau-Brunswick",
  "Quebec - Québec", "Ontario", "Manitoba", "Saskatchewan", "Alberta",
  "British Columbia - Colombie-Britannique"
)

res <- ds |>
  filter(
    geo_name %in% provinces,
    profile  %in% c("Mother tongue - English", "Mother tongue - French",
                    "Mother tongue - Other")
  ) |>
  select(geo_name, profile, value) |>
  collect()
```

That query returns 30 rows — the real decoded 1981 counts.

## 4. Work with the data

A little tidying — short province labels, and mother-tongue shares within each
province — then a stacked bar chart. In 1981 Quebec is overwhelmingly
French-mother-tongue while New Brunswick is the one clearly bilingual province:

```{r plot, fig.width = 7, fig.height = 4.2, fig.alt = "Stacked bar chart of 1981 mother-tongue composition by province: English, French and Other shares, ordered by French share. Quebec is overwhelmingly French, New Brunswick roughly one-third French, the rest predominantly English."}
library(ggplot2)

plot_data <- res |>
  mutate(province = sub(" - .*$", "", geo_name), # shorter province name
         mt = factor(sub("Mother tongue - ", "", profile),
                     levels = c("French", "Other", "English"))) |>
  mutate(share = value / sum(value), .by = province)

# order provinces by their French share
french_order <- plot_data |>
  filter(mt == "French") |>
  arrange(share) |>
  pull(province)

plot_data |>
  mutate(province = factor(province, levels = french_order)) |>
  ggplot(aes(share, province, fill = mt)) +
  geom_col(width = 0.75) +
  scale_x_continuous(labels = function(x) paste0(x * 100, "%"),
                     expand = expansion(mult = c(0, 0.03))) +
  scale_fill_manual(NULL,
    values = c(French = "#2c7fb8", Other = "#c7c7c7", English = "#d95f0e"),
    breaks = c("English", "French", "Other")) +
  labs(y = NULL, x = "Share of population",
       title = "Mother tongue by province, 1981 Census",
       caption = "Data: Decoded from the Beyond 20/20 .ivt file (97-570-X1981004)")
```

## A note on cells that aren't there

An IVT stores only the cells that have a value, so a coordinate missing from the
result is **not** necessarily a zero — it may be a suppressed or not-applicable
value. Which it is, the file does say, in a status block after each page's
values; it is off by default because on a large sparse crosstab those cells
outnumber the data. Turn it on when you intend to fill the grid in:

```r
tab <- ivt_read_table("98100040", missing = TRUE)

ivt_tidy(tab, missing = TRUE)          # missing cells as `value = NA` rows,
                                       # with `symbol` / `status` saying why
ivt_missing(tab)                       # just those cells
attr(ivt_missing(tab), "legend")       # the symbols, in the file's own words

get_statcan_ivt("98-10-0040", missing = TRUE)   # cached alongside the data
```

Without it, completing the grid with zeros turns every suppressed value into a
`0`. See `?read_ivt` for the caveats (a few vintages write no status block at
all, and each gap warns rather than being folded into the count).

## Recap

```r
library(canivt); library(dplyr); library(ggplot2)

statcan_ivt_catalogue() |>                       # 1. find the table
  filter(census_year == 1981, grepl("Profile", title))

ds  <- get_statcan_ivt("97-570-X1981004")        # 2. download + decode + cache
res <- ds |> filter(...) |> collect()            # 3. query the slice you want
ggplot(res, ...) + geom_col()                    # 4. graph
```

The same four steps work for **any** table in the catalogue — 1981 profiles,
2021 dissemination-area crosstabs, commuting-flow tables — because `read_ivt()`
(which `get_statcan_ivt()` calls) decodes every vintage through one path. For the
full public API see `?read_ivt` and `?get_statcan_ivt`; for the binary format
itself see `vignette("ivt-format")`.
