---
title: "Getting started: reading an older census table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: reading an older census table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette walks through a complete, realistic use case: **look up a table in
the Statistics Canada catalogue, download and decode it, and draw a graph** — a
handful of lines end to end.

We deliberately pick an **old** table: the **1981 Census Profile for Canada,
Provinces, Territories, Census Divisions and Census Subdivisions**
(`97-570-X1981004`). Modern census tables come with a tidy CSV and a Web Data
Service API, but the pre-2001 vintages are published **only** as *Beyond 20/20*
`.ivt` files — there is no CSV, no API, no easy-to-parse alternative. `canivt`
decodes them straight from the bytes, so a 1981 profile becomes an ordinary data
frame.

```{r setup, message = FALSE}
library(canivt)
```

## (Optional) set a cache

`canivt` can cache downloaded `.ivt` files and the decoded Parquet so the second
run is instant. Point it at a directory (e.g. in `.Renviron` via
`CANIVT_IVT_CACHE` / `CANIVT_DATA_CACHE`, or per session):

```{r cache, eval = FALSE}
options(
  canivt.ivt_cache  = "~/canivt-cache/ivt",   # raw .ivt downloads
  canivt.data_cache = "~/canivt-cache/data"   # decoded Parquet + catalogue
)
```

Without a cache everything still works; it just re-downloads each session.

## 1. Look up the table in the catalogue

`statcan_ivt_catalogue()` scrapes the StatCan census datasets index into one
tibble (one row per published table version), cached as Parquet after the first
call. Each row carries the `catalogue` number, `census_year`, `title`, `topic`
and the download URLs.

```{r catalogue, eval = FALSE}
catl <- statcan_ivt_catalogue()

# Find the 1981 census-division / subdivision profile
library(dplyr)
catl |>
  filter(census_year == 1981, grepl("Profile.*Census Divisions", title)) |>
  select(catalogue, census_year, title)
#> # A tibble: 1 × 3
#>   catalogue        census_year title
#>   <chr>                  <int> <chr>
#> 1 97-570-X1981004         1981 Census Profile for Canada, Provinces and Terr…
```

If you already know the product id you can skip this step entirely and go
straight to step 2.

## 2. Download and decode with `get_statcan_ivt()`

`get_statcan_ivt()` resolves the id against the catalogue, downloads the `.ivt`,
decodes it, caches the result as Parquet, and hands back an
[`arrow`](https://arrow.apache.org/docs/r/) dataset connection. The second call
skips the download and decode.

```{r get, eval = FALSE}
ds <- get_statcan_ivt("97-570-X1981004")
ds
#> FileSystemDataset with 1 Parquet file
#> geo_label: string
#> geo_name: string
#> geo_uid: string
#> values: string
#> profile: string
#> value: double
```

The table is a classic profile: one row per **geography × profile
characteristic**, with the count in `value`. The `profile` column holds the ~80
characteristics (population, age groups, mother tongue, tenure, …) and `geo_name`
the bilingual place name.

```{r explore, eval = FALSE}
ds |> distinct(profile) |> collect() |> head(6)
#> # A tibble: 6 × 1
#>   profile
#>   <chr>
#> 1 Population, 1976
#> 2 Population, 1981
#> 3 Land area in square kilometres, 1981
#> 4 Population density per square kilometre, 1981
#> 5 Total population by sex and age
#> 6 Total male population
```

## 3. Query the data you want

Because `ds` is an Arrow dataset you can filter and select with `dplyr` verbs and
only `collect()` the rows you need. Here we pull the **mother-tongue** counts
(English / French / Other) for the ten provinces:

```{r query, eval = FALSE}
provinces <- c(
  "Newfoundland - Terre-Neuve", "Prince Edward Island - Île-du-Prince-Édouard",
  "Nova Scotia - Nouvelle-Écosse", "New Brunswick - Nouveau-Brunswick",
  "Quebec - Québec", "Ontario", "Manitoba", "Saskatchewan", "Alberta",
  "British Columbia - Colombie-Britannique"
)

res <- ds |>
  filter(
    geo_name %in% provinces,
    profile  %in% c("Mother tongue - English", "Mother tongue - French",
                    "Mother tongue - Other")
  ) |>
  select(geo_name, profile, value) |>
  collect()
```

That query returns 30 rows — the real decoded 1981 counts. For a self-contained
example, here they are inline:

```{r data, echo = FALSE}
res <- data.frame(
  geo_name = rep(c("Alberta", "British Columbia - Colombie-Britannique",
                   "Manitoba", "New Brunswick - Nouveau-Brunswick",
                   "Newfoundland - Terre-Neuve", "Nova Scotia - Nouvelle-Écosse",
                   "Ontario", "Prince Edward Island - Île-du-Prince-Édouard",
                   "Quebec - Québec", "Saskatchewan"), each = 3),
  profile = rep(c("Mother tongue - English", "Mother tongue - French",
                  "Mother tongue - Other"), times = 10),
  value = c(1810545, 62150, 365030, 2249310, 45615, 449540, 735920, 52555,
            237760, 453310, 234035, 9060, 560465, 2655, 4565, 793165, 36030,
            18250, 6678765, 475605, 1470735, 115045, 6080, 1380, 706110,
            5307010, 425275, 770815, 25540, 171960)
)
```

## 4. Draw the graph

A little tidying — short province labels, and mother-tongue shares within each
province — then a stacked bar chart. In 1981 Quebec is overwhelmingly
French-mother-tongue while New Brunswick is the one clearly bilingual province:

```{r plot, fig.width = 7, fig.height = 4.2, fig.alt = "Stacked bar chart of 1981 mother-tongue composition by province: English, French and Other shares, ordered by French share. Quebec is overwhelmingly French, New Brunswick roughly one-third French, the rest predominantly English."}
library(ggplot2)

res$province <- sub(" - .*$", "", res$geo_name)                 # English short name
res$mt       <- factor(sub("Mother tongue - ", "", res$profile),
                       levels = c("French", "Other", "English"))

# share of each mother tongue within the province
res$share <- ave(res$value, res$province, FUN = function(v) v / sum(v))

# order provinces by French share
fr <- res$share[res$mt == "French"]
res$province <- factor(res$province,
                       levels = res$province[res$mt == "French"][order(fr)])

ggplot(res, aes(share, province, fill = mt)) +
  geom_col(width = 0.75) +
  scale_x_continuous("share of population",
                     labels = function(x) paste0(x * 100, "%"),
                     expand = expansion(mult = c(0, 0.03))) +
  scale_fill_manual(NULL,
    values = c(French = "#2c7fb8", Other = "#c7c7c7", English = "#d95f0e"),
    breaks = c("English", "French", "Other")) +
  labs(y = NULL, title = "Mother tongue by province, 1981 Census",
       subtitle = "Decoded from the Beyond 20/20 .ivt file (97-570-X1981004)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(), legend.position = "top")
```

## Recap

```r
library(canivt); library(dplyr); library(ggplot2)

statcan_ivt_catalogue() |>                       # 1. find the table
  filter(census_year == 1981, grepl("Profile", title))

ds  <- get_statcan_ivt("97-570-X1981004")        # 2. download + decode + cache
res <- ds |> filter(...) |> collect()            # 3. query the slice you want
ggplot(res, ...) + geom_col()                    # 4. graph
```

The same four steps work for **any** table in the catalogue — 1981 profiles,
2021 dissemination-area crosstabs, commuting-flow tables — because `read_ivt()`
(which `get_statcan_ivt()` calls) decodes every vintage through one path. For the
full public API see `?read_ivt` and `?get_statcan_ivt`; for the binary format
itself see `vignette("ivt-format")`.
