---
title: "The Beyond 20/20 IVT file format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Beyond 20/20 IVT file format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = nzchar(Sys.getenv("COMPILE_VIG")))
```

This vignette documents the Statistics Canada *Beyond 20/20* `.ivt` binary
format as reverse-engineered for `canivt`, and shows how the package's functions
map onto each part of the file. It is aimed at anyone who wants to understand,
validate, or extend the parser. For day-to-day use see `?read_ivt`. Two more detailed references ship with the
package: the authoritative byte-format spec
`system.file("notes/ivt-format.md", package = "canivt")`, and a terse catalog of
every byte marker/signature the decoder keys on,
`system.file("notes/markers.md", package = "canivt")` (self-checked by
`tests/testthat/test-markers.R`).

## One layout, one decoder

Every `.ivt` starts with a 4-byte signature ending `00 20 00`; the modern
census/custom lineage sets the leading byte to `04`. (A second, older container
generation sets it to `02` instead — see "An older survey-generation container"
near the end of this vignette; everything in this section applies to both.) The
file is one contiguous binary, and there is a single, descriptor-driven,
name/type-agnostic decoder: it
reads the dimension structure from the header, nests **every** dimension into a
power-of-two positional bitmap (data dimensions innermost, geography outermost),
and walks the value pages. What used to be documented as "family 1" and
"family 2" are **not two formats** — they are two cases of this one layout,
differing only in *which* dimension straddles the fixed **2048-bit (256-byte)
page boundary**:

* **A data dimension straddles** → geography is pushed fully into the page
  directory, one page per (geography, outer-data-coordinate). The reference table
  is **98-10-0241** (housing indicators by tenure; 166 geographies, 7 dimensions).
  Historically "family 1."
* **The data dimensions fit in ≤ 2048 bits** → *geography* straddles: several
  geographies share each page's presence record, and the directory is a flat list
  of geography-window pages. The reference table is **98-10-0023** (age × gender,
  down to dissemination areas; 63,404 geographies, 14.5 M cells). Historically
  "family 2."

The **1991 census** files (e.g. **1003011**, E9101 — population by single year of
age × sex) are the *same* container with pre-DGUID conventions: `int16`/`int32`
value pages and a bilingual, inline codebook. They decode through the same path.

`read_ivt()` auto-detects the case via `ivt_family()` (which just reports whether
geography straddles), but the cell decode and metadata read are **shared for
every file** — `family` only tags provenance. Every `.ivt` in the test corpus
decodes, and no file in it is currently refused. That is not a promise about
files the corpus has never seen: detection is structural, and `read_ivt()` still
refuses anything whose page geometry it cannot verify rather than emit misindexed
cells.

```{r}
library(canivt)
tab <- read_ivt(ivt_download("98100023"))
tab
```

## File layout at a glance

The high-level region order is the same for every IVT: a small **header**, the
**page directory**, the **value pages** (which dominate — typically ~85–90 % of
the file), then the **codebook** (geography identifiers/labels and the data
dimension members), with **footnotes** near the tail.

```{r layout-plot, eval = TRUE, fig.width = 7, fig.height = 3.6, message = FALSE, fig.alt = "Waffle chart of the IVT byte layout on a 20-by-10 grid of 200 tiles filled in file order: header, page directory, the dominant value-pages region (~175 tiles), the geography codebook (~24 tiles), footnotes and dimension blocks. Each region's tile count is proportional to its byte size, with a one-tile minimum so the slivers stay visible."}
library(ggplot2)
library(dplyr)

# Representative byte regions of the geography-straddle reference table
# 98-10-0023 (142,016,485 bytes), in file order. Every IVT follows this order.
regions <- data.frame(
  region = c("Header", "Page directory", "Value pages (data)",
             "Geography codebook", "Footnotes", "Dimension blocks (Age/Gender)"),
  size   = c(35950, 126800, 124122962, 17603000, 3000, 120485)
)

# Allocate a 20x10 = 200-tile waffle grid: each region gets tiles proportional
# to its byte size, but at least one tile so even the slivers show. The floor
# replaces the old ad-hoc per-region "exaggeration" scaling.
n_cols <- 20L
n_rows <- 10L
n_tiles <- n_cols * n_rows

alloc_tiles <- function(sizes, total, min_tiles = 1L) {
  tiles <- rep(min_tiles, length(sizes))     # everyone starts at the floor
  extra <- total - sum(tiles)                # tiles left to hand out
  ideal <- extra * sizes / sum(sizes)        # proportional, largest-remainder
  base  <- floor(ideal)
  tiles <- tiles + base
  short <- extra - sum(base)                 # rounding shortfall
  if (short > 0) {
    take <- order(ideal - base, decreasing = TRUE)[seq_len(short)]
    tiles[take] <- tiles[take] + 1L
  }
  tiles
}

regions$tiles <- alloc_tiles(regions$size, n_tiles)

# One row per tile, filled in file order, then wrapped onto the grid reading
# left-to-right, top-to-bottom.
regions$region <- factor(regions$region, levels = regions$region)
tile_region <- rep(regions$region, regions$tiles)
idx <- seq_along(tile_region) - 1L
tiles <- data.frame(
  region = tile_region,
  col    = idx %%  n_cols,
  row    = idx %/% n_cols
)

# legend label: real size plus the tile count it maps to
human <- function(b) ifelse(b >= 1e6, paste0(round(b / 1e6, 1), " MB"),
                     ifelse(b >= 1e3, paste0(round(b / 1e3),  " KB"),
                                      paste0(b, " B")))
labels <- paste0(regions$region, " — ", human(regions$size))

ggplot(tiles, aes(x = col, y = row, fill = region)) +
  geom_tile(colour = "white", linewidth = 0.8, width = 0.95, height = 0.95) +
  scale_y_reverse(NULL, breaks = NULL, expand = c(0, 0)) +
  scale_x_continuous(NULL, breaks = NULL, expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set2", name = NULL, labels = labels) +
  coord_equal() +
  labs(title = "IVT file layout — Table 98-10-0023 (142 MB)",
       subtitle = "One tile ≈ 0.5 % of the file; regions fill in file order (≥1 tile each)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right", panel.grid = element_blank(),
        plot.title.position = "plot")
```

So roughly 87 % of the file is data; the geography codebook is the second-largest
region (~12 %), and the header, directory, footnotes and per-dimension member
blocks are slivers. All multi-byte integers are **little-endian**, and value runs
frequently begin on **odd (unaligned) byte offsets** — a property that defeats
naive aligned scans. Byte offsets in the parser are 0-based to match the layout.

## Reading the layout from the header

The fixed header points at every major section, so the layout can be read up
front without scanning for markers. `ivt_f2_header_layout()` (uniform across the
modern and legacy formats) returns these as byte offsets:

| header offset | section |
|---|---|
| `u32 @32` | dimension descriptor (per-dimension count, type, name + title) |
| `u32 @40` / `u32 @48` | French / English title blocks — **set in the legacy format, zero in the modern one** (the version indicator) |
| `u32 @552` | geography field/attribute count (11 modern / 12 legacy) |
| `u16 @558` | page-directory start — **low 16 bits only** (the true start is `u16 + k·65536` for the smallest `k` whose first entry validates; `k = 0` for the reference tables, `k > 0` for a few large tables) |
| `u32 @572` | codebook region start |

A further **section-pointer region** (`≈ @544..1080`) is read the same way, each
pointer resolving to a block directory of the identical 8-byte entry shape: `@544`
→ the master directory (titles, the dimension descriptor, identity/notes blobs,
product id), `@712` → the data-quality-flag legend, and `@824 + 14·(k−1)` → the
per-dimension codebook block directory (member labels, ordinals, footnotes) for
dimension `k`. So the codebook, notes and legends are all located **from the
header**, with bounded tail scans surviving only as fallbacks.

The page directory itself is a contiguous run of `[u32 offset][u16 size][u16 size]`
records (the two sizes agree; the size is the page's *allocated* length), in
geography member order. Each record points at a *page* of value data.

## Dimensions

The dimension descriptor declares each dimension uniformly. For 98-10-0023:
Geography (63,404) × Age (128) × Gender (3). The decoder nests them
**positionally** — geography outermost, the data dimensions inner in descriptor
order with the **last** (here Gender) fastest-varying / the unit of value storage
— and never asks which dimension is which by name or type byte. Geography is the
first descriptor dimension in every layout **except the profile lineage** (a few
1981/1991 profile tables store a one-member "Values" placeholder first and put
geography *last*); the nesting handles that with no special case, because it is
purely positional. Exactly one dimension straddles the 2048-bit page boundary
(§ below), which is what the historical "family" label really named.

```{r}
meta <- ivt_metadata(ivt_download("98100023"))
data.frame(dimension = meta$dimension_names,
           members   = meta$dimension_counts)
```

### Three witnesses to the member count

Getting a dimension's member count wrong misindexes every cell downstream, and
the descriptor's count field alone is not always trustworthy — its record framing
is ambiguous in places, and a count can be read at the wrong width. The file
states the same number three times, in three independent places, so the count is
**reconciled** rather than taken on faith:

1. the **descriptor** record's count field;
2. the **codebook** — the declared slot allocation and the length of the member
   arrays themselves. Member labels are written in blocks capped at 256, so a
   dimension with more members spans several blocks, and the run of block lengths
   witnesses the true total;
3. the **page directory** — how far the outermost paged dimension's entries
   actually extend, counting only entries that both decode and carry cells.

Reconciliation only ever *raises* a count, never lowers one, and every
correction is a loud, classed warning. When the count is right the container
agrees exactly: the directory's outer entry cartesian equals the page count.

## Case A — geography straddles: a single page directory (98-10-0023)

When the data dimensions fit in one 2048-bit presence record, geography is the
dimension that straddles the page boundary, so several geographies share each
page and the directory is one flat list. Here each directory record is a page
that packs **4 geographies**:

```
[4-byte marker][4 × 64-byte presence records][0xFF trailer][head block][dense value run]
```

* the **marker**'s low nibble is the value width: `0x8` → float64, `0x4` →
  int32, `0x2` → int16 (so age counts are int16 but the average/median-age
  statistics are float64); its third byte encodes the trailer length and its
  fourth a head block of `32·(b3 − 8)` bytes (0 on most pages). Trailer and head
  together are not padding: they are the **index bitmap** for the page's trailing
  cell-status block (below), which is why the head grows exactly on the vintages
  that write a big one;
* each geography's **presence** is a 64-byte record stored **byte-pair-swapped**
  — swap adjacent bytes, then read a positional nibble per member, with the three
  genders Total/Men/Women at bits 3/2/1 (the all-present marker is `0xE`);
* **values** are dense little-endian over the present `(member, gender)` cells,
  member-major / gender-inner. Only non-zero cells are stored — the StatCan CSV
  publishes the zeros, so an absent cell is *either* a zero or a missing value.

Which of the two it is is written down. Every page carries a *cell-status block*
after the value run, and it is what distinguishes an absent-but-zero cell from an
absent-and-missing one — so the familiar "absent means 0" rule is a
simplification, in every vintage, not just the modern ones.

On pages whose marker byte has high nibble `0xa` the block is a self-describing
array of `W`-bit codes, one per cell, whose two-bit vocabulary is `0` (a value,
or a genuine zero), `1` (filler), `2` (`x`, suppressed for confidentiality) and
`3` (`...`, not available) — validated cell-exact against StatCan's published
tables on 98-10-0040, 98-10-0128, 98-10-0655 and 98-10-0658.

Its **addressing** is not yet general (three sampled tables index it three
different ways), so `canivt` counts these pages and reads no further rather than
guessing.

On high-nibble `0x8` pages — which is every pre-2016 vintage — the block is a
bare one-bit-per-cell mask, and it says the same thing more crudely: it flags a
**subset** of the absent cells, and that subset is the genuine zeros. The
unflagged remainder is the missing data (the Beyond 20/20 viewer prints `N`).
Where a table has no missing values the mask coincides with the complement of the
presence bitmap, which is the usual case and the reason it long looked
redundant. Verified against the viewer on the 2001 census profile
`97F0020XCB2001070`.

**This mask is decoded**, opt-in:

```r
x <- read_ivt(path, missing = TRUE)
x$missing        # one row per cell the file says is MISSING, same coordinate
                 # columns as x$cells, minus the value
```

The mask is stored as a sparse array of value-width words, addressed by the index
bitmap in the trailer + head region; the gate is the length identity
`popcount(index) · width == tail length`, which holds on all 1,342,037 mask pages
of the development corpus. It is off by default because completeness is
vintage-dependent: some lineages write no tail at all, `0xa` pages are not
decoded, and a mask that stops short of the grid leaves the cells past it
unmasked for want of a word rather than by the file's statement. Each of those
raises its own classed warning (`canivt_status_unreadable`,
`canivt_status_block_undecoded`, `canivt_status_beyond_mask`, …), so a missing
count always comes with the caveats attached rather than hidden.

Whole-geography suppression is surfaced regardless, via
`metadata$geographies$has_data`. See `inst/notes/ivt-format.md` for the byte
layout, the second (undecoded) tail block, and the open `0xa` addressing problem.

The full per-geography codebook — name, DGUID, geographic level and type
(+ abbreviation), two geocodes, the data-quality flag and note, and the
non-response rate — decodes exactly for all 63,404 geographies; it is attached by
`read_ivt(geo_attributes = TRUE)` (internally `ivt_f2_geo_attributes()`), while
the default metadata path already carries the names and DGUIDs.

## Case B — a data dimension straddles: per-geography directories (98-10-0241)

When the data dimensions alone overflow 2048 bits, a data dimension straddles
instead and geography is pushed fully into the directory — **each geography gets
its own page directory** at a fixed stride:

```
[0]            header: identity + dimension names
[37167]        geography index: one page-directory per geography, stride 0x1000
~1.08M..~55.4M value pages
~56.93M..EOF   footnote legend + codebook
```

> **Gotcha.** The directories are grouped 8 to a `0x8000` region (288 entries
> used per `0x1000` slot, the rest zero-padded). Striding by `0x8000` instead of
> `0x1000` silently reads only every 8th geography — 21 of 166. The correct
> stride is `0x1000`.

Here the page presence is a **256-byte positional, dimension-padded** bitmap:
32-byte rows per Period; Statistics in 8-byte blocks; Housing indicators 0..5
within a block (the two bytes of each adjacent housing pair are **byte-swapped**,
so housing `h` is read at `stat*8 + bitwXor(h, 1)`); and each presence byte's
bits 7..1 flag the 7 Tenure members (bit 0 padding). `0xFE` therefore means "all
7 tenure values present".

```{r, eval = TRUE}
# bits 7..1 of 0xFE flag the 7 tenure members; bit 0 is padding
bitwAnd(bitwShiftR(0xFE, 7:1), 1L)
#> [1] 1 1 1 1 1 1 1
```

This is the same presence machinery as Case A, just with a data dimension in the
straddle role: the presence granularity is the **innermost** dimension, the "all
present" marker is `2^n − 2` over that bit width, the per-dimension strides are
power-of-two padded and fixed, and one byte of each adjacent pair is swapped. The
single decoder (`ivt_layout()` + `ivt_decode()`) computes the straddle dimension
and these strides from the descriptor, so both cases run the identical code.

The padding itself is **declared metadata**, not a derived rule: every
dimension's codebook directory carries a member block (`81 02 <alloc-u16> 16 00`,
or the survey generations' time-series table `... 08 00`) whose leading u16 is
the dimension's allocated member-slot capacity, and every nesting level —
presence bits and directory strides alike — is padded to that allocation. It
equals `nextpow2(count)` on almost every table, but not always: one Labour Force
Historical Review table allocates 32 slots for a 10-member dimension, and its
page directory faithfully spans the larger cartesian (the padding slots hold
minimal, empty pages). Reading the declared allocation instead of re-deriving
`nextpow2` is what lets one layout rule cover that lineage too.

That block declares more than the capacity. Its mid-section is a table of
**22 bits per allocated slot** (pair-swapped, MSB-first): a *live* bit, the
member code's length in unary, and a flag for an extra trailing code byte. So the
file states **which slots hold members and which were deleted** — the bitmap
addresses members by *slot*, and slots can have holes. The declaration is trusted
only when walking the code lengths it predicts consumes the following member-code
array byte-exactly; where it does, the count and slot positions are read rather
than guessed. The same slot map also addresses the **codebook** arrays, which may
be written one record per allocated slot and left empty at the rest — an interior
hole otherwise defeats a trailing-NA trim and silently shifts every label past it.

## The codebook: labels, geographic ids, footnotes

The tail of the file holds the codebook: for each dimension, several parallel,
**member-ordered** arrays of length-prefixed ("Pascal": one length byte then that
many text bytes) strings — member ordinals, the name (English then French) and,
for Geography, the level, abbreviations, classification code, and full **DGUID**
(`2021A000011124` = Canada). Labels are encoded in **Windows-1252**: the
`0x80–0x9F` block carries real punctuation (e.g. `0x92` = the curly apostrophe in
`Tla'amin Lands`), so the byte-class test that frames Pascal records must accept
it — otherwise such a label aborts a record and splits the array mid-stream.

```{r}
# labelled long table; read_ivt(geo_attributes = TRUE) attaches the full
# geography attribute table so tidy can label by name + level.
tab <- read_ivt(ivt_download("98100023"), geo_attributes = TRUE)
ivt_tidy(tab)
```

Footnotes are read **from the header**, not by scanning. Table-level (cube) notes
come from the master directory; each dimension's notes come from its per-dimension
slot directory (the `@824 + 14·(k−1)` table above). Within a dimension each note
carries a **scope** — a *member* note (annotating one geography or data member) or
a whole-*dimension* note — resolved structurally: a `84 01` member bitmap opens the
footnote region and lists, in member order, which members carry a note, so each
note surfaces with its `scope`, owning `dimension` and `member_id`/`member_refs`.
The text itself is tagged `Footnote N` (EN) / `Renvoi N` (FR); a bounded tail scan
of those markers survives only as a loud fallback when the slot directories list
none. (Those same `Renvoi N` texts also appear per-member in the geography
directory's tail as `[01 01][u16 len-4][01]` text blobs — told apart from member
arrays by their NUL-free payload so they are not miscounted as attributes; see
`markers.md` §F.) The legacy pre-DGUID files instead cite notes as `(N)` markers
embedded in the member labels, linked back to the members that carry them.

```{r}
str(meta$footnotes[[1]])
length(meta$footnotes)
```

## The 1991 (pre-DGUID) files (1003011)

The 1991 census files are the *same* container in the geography-straddle case,
with pre-DGUID conventions — fully decoded through the same `read_ivt()` path
(cell-exact against the scraped Beyond 20/20 ground truth):

* value pages are `int16`/`int32` (markers `0x82`/`0x84`) rather than float64,
  and presence uses the same **byte-pair-swapped** positional bitmap;
* geography is a single inline block per chunk of `"<name> (<GEOUID>) <flag>"` —
  a **bilingual** name (`Newfoundland | Terre-Neuve`), a bare **GEOUID** (a
  shortened DGUID without the year/area-type prefix), and a data-quality flag;
* the table identity is stored *out of line* (the `@40`/`@48` header pointers,
  which are zero on the modern DGUID files — that difference is the format-version
  signal), and footnotes are the `(N) text` form described above.

A profile lineage of pre-DGUID files (some 1981/1991 profiles) additionally puts
geography *last* behind a one-member "Values" placeholder; the positional nesting
absorbs that without a special case.

Pre-2016 tables predate the modern 8-digit b2020 `.zip` endpoint that
`ivt_download()` uses, so fetch the raw `.ivt` through its catalogue row
(`statcan_ivt_download()` resolves the `Download.cfm?PID=` link and returns the
local path), then read it the same way:

```{r}
row <- subset(statcan_ivt_catalogue(), catalogue == "1003011")
leg <- read_ivt(statcan_ivt_download(row))   # 41,859 geographies, Age(110) × Sex(3)
ivt_tidy(leg)
```

## An older survey-generation container (`02 00 20 00`)

A second, older Beyond 20/20 container generation shares the same page/value/
codebook model but starts with `02 00 20 00` instead of `04 00 20 00` — byte 0
is a container-generation tag, not part of a fixed constant, and
`ivt_family()`/`read_ivt()` handle both transparently. This generation covers
older survey product lines rather than census geography tables: Health
Statistics at a Glance (1999), the 1996 Census of Agriculture, the 1996
Small Area Business survey, and the provincial Canadian Business Patterns
releases of 1997–2002 (see "Undeclared geometry" below).

Two things are structurally different, both confined to the descriptor and
codebook:

* **No geography dimension.** These are single-area survey products — the
  `REGION`/`GEOGRAPHY` dimension carries no DGUID or geographic identifiers, so
  `read_ivt()` treats it as an ordinary, fully-labelled data dimension:
  `metadata$geographies` is empty and `cells` has no `geo` column.
* **Generated time-series labels.** A reference/time dimension can be stored
  with no code or label array at all — only a per-member **date table**
  (one-byte populated-slot flags + a 3-byte date, days since `0000-03-01`).
  Member labels (typically a year) are *generated* from those dates. The slot
  addressing is the same one described above — this table (`... 08 00`) is simply
  where a dimension declares it when it carries no `16 00` member-code block. A
  dimension carries one or the other, never both.

```{r, eval = FALSE}
# Health Statistics at a Glance 1999 is Borealis-hosted (see vignette("borealis")
# for browsing/downloading): Quantifier x Geography x Period, no geography
# dimension, year labels generated from the on-disk dates.
path <- borealis_ivt_download(hits[1, ])   # hits from borealis_ivt_catalogue()
hsg  <- read_ivt(path)
ivt_tidy(hsg)
```

Values in this generation are complete integers in the indicator's own unit —
not a fixed-point encoding needing a decimal scale. The unit is stated in the
relevant member's description text, surfaced via `ivt_members()`'s
`description`/`description_fr` columns where it can be mapped unambiguously.

See `inst/notes/ivt-format.md` ("The older `02 00 20 00` survey-generation
container") and `inst/notes/markers.md` §E.1 for the exact byte framings.

### Undeclared geometry: the provincial Business Patterns cluster

One corner of this generation declares *less* than everything above assumes. The
provincial Canadian Business Patterns releases (`PROVSIC*`, `PRVNAIC*`,
`CACMA*`, 1997–2002; province/CMA × industry × employment-size class) carry **no
`16 00` slot table for any dimension**, so the outer directory stride — the
number of entry slots each geography occupies — is stated nowhere in the file and
has to be *measured* from the page directory itself. It is measured as a
**tiling**, not a progression: every geography occupies `S` consecutive entry
slots and writes the same window residues inside them, and the smallest `S` that
tiles the whole directory wins. That matters because a run need not start at
window 0 — one file lays its 11 industry windows at slots 3..13 of a 16-slot
group. A written page whose presence record is entirely zero is treated as an
**absence**: it carries no cells, so it witnesses nothing about where members sit.

Because the geometry is inferred rather than declared, it is **adopted only if
the decoded values reconcile exactly** on the file's own arithmetic — the
industry `Total` member equalling the sum of the detail members for every
(geography, size class), or `Canada` equalling the sum of the provinces. A file
whose geometry cannot be measured, or whose decode does not reconcile, is left
**unsupported** rather than emitting unvalidated numbers.

Two caveats travel with these tables, both raised as loud, classed warnings:

* the industry **labels are provisional** (`canivt_suba_labels`) — reconciliation
  validates the *sums*, not the code → member assignment, and a uniform relabel
  would leave every sum unchanged. There is no published ground truth to settle
  it against;
* the recovered geometry itself is flagged (`canivt_suba`), so a caller can spot
  these tables programmatically or promote the warnings to errors with
  `options(canivt.strict = TRUE)`.

Values and dimension structure are otherwise ordinary. See `R/suba.R` and the
sub-A sections of `inst/notes/unsupported-formats.md` and
`inst/notes/coverage.md`.

## How `canivt` maps to the format

| Region | Function(s) | Source file(s) |
|--------|-------------|----------------|
| Header identity, dimension descriptor, layout | `ivt_metadata()`; (internal) `ivt_f2_header_layout()`, `ivt_f2_descriptor()` | `R/codebook-f2.R`, `R/dimdir.R` |
| Page directory / geography index | (internal) `ivt_layout()` | `R/decode.R`, `R/container*.R` |
| Presence bitmap + value codec (both straddle cases) | (internal) `ivt_decode()` | `R/decode.R`, `R/decode-f2.R` |
| Codebook labels, DGUIDs/GEOUIDs, geography attributes, footnotes | `ivt_metadata()`, `read_ivt(geo_attributes = TRUE)`; (internal) `ivt_f2_geo_attributes()` | `R/codebook-f2.R`, `R/dimdir.R`, `R/read-f2.R` |
| Tidy / write outputs | `ivt_tidy()`, `ivt_write_*()` | `R/read.R`, `R/write.R` |

## Validation

`canivt` is validated against the StatCan CSV/metadata downloads: the
data-dimension-straddle reference (98-10-0241) all 166 geographies and 7,489,464
cells exact; the geography-straddle reference (98-10-0023) all 63,404 geographies
and 14.5 M cells exact, plus every geography attribute and the Age/Gender labels;
and the 1991 table (1003011) cell-exact for all scraped ground-truth geographies.
The unified decoder is additionally **byte-identical** to the two former decoders
on six reference tables, and viewer/CSV-validated across the wider corpus —
1996–2021 census tables, 1981/1991 profiles, 2001/2006 F-series, large 2016
`98-400-X` crosstabs, commuting-flow tables, custom extracts, the Business
Register lineage, and the older `02 00 20 00` survey generation. An opt-in
regression ledger runs the whole local corpus through `read_ivt()` and asserts
the exact cell count per table. For older vintages the package scrapes the
Beyond 20/20 web viewer for ground-truth data to validate against. See the
`tests/` directory and the notes file
(`inst/notes/coverage.md`, `decode-history.md`) for the exact figures.
