---
title: "The Beyond 20/20 IVT file format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Beyond 20/20 IVT file format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette documents the Statistics Canada *Beyond 20/20* `.ivt` binary
format as reverse-engineered for `canivt`, and shows how the package's functions
map onto each part of the file. It is aimed at anyone who wants to understand,
validate, or extend the parser. For day-to-day use see `?read_ivt`.

The running example is StatCan table **98-10-0241** ("Housing indicators by
tenure … by household type and period of construction"), whose `.ivt` is the
reference the parser is validated against.

```{r}
library(canivt)
path <- ivt_download("98100241")
tab  <- read_ivt(path)
tab
```

## Dimensions

The table has 7 dimensions. The first, **Geography**, indexes the data blocks;
the remaining six are the "inner" dimensions of every data page. The innermost
dimension, **Tenure** (7 members), is the unit of value storage.

| # | Dimension | Members |
|---|-----------|--------:|
| 1 | Geography | 166 |
| 2 | Age of primary household maintainer | 9 |
| 3 | Household type incl. census family structure | 16 |
| 4 | Period of construction | 13 |
| 5 | Statistics | 3 |
| 6 | Housing indicators | 6 |
| 7 | Tenure incl. mortgage/subsidy | 7 |

That is `9 × 16 × 13 × 3 × 6 × 7 = 235,872` cells per geography. Cells that are
zero are not stored, so the decoded table is sparse (~7.49 M non-zero cells
across all 166 geographies).

## File layout

An `.ivt` file is one contiguous binary with five regions:

```
[0]            header: table identity + dimension names (starts 04 00 20 00)
[37167]        geography index: one page-directory per geography
~1.08M..~55.4M page data: the value pages (~98% of the file)
~56.93M        footnote legend (EN "Footnote N", FR "Renvoi N")
~56.94M..EOF   codebook: member labels + geography DGUIDs
```

All multi-byte integers are **little-endian**, and value runs frequently begin on
**odd (unaligned) byte offsets** — a property that defeats naive aligned scans.
Throughout the parser, byte offsets are 0-based to match the layout; the helpers
in `R/utils-bytes.R` convert to R's 1-based indexing.

## The geography index

Per-geography *page directories* begin at byte `37167` with a stride of
`0x1000` (4096 bytes). Directory `n` (0-based) is geography **Member ID `n + 1`**
in the StatCan metadata — i.e. geographies are in member order.

> **Gotcha.** The directories are grouped 8 to a `0x8000` region (288 entries
> used per `0x1000` slot, the rest zero-padded). Striding by `0x8000` instead of
> `0x1000` silently reads only every 8th geography — 21 of 166. This was the
> original bug; the correct stride is `0x1000`.

Each directory holds up to 288 entries of `[u32 offset][u16 size][u16 size]`
(the two size fields are equal; an entry is valid when they agree and the offset
lands in the data region). The 288 entries are **positional** over the three
outermost inner dimensions:

```
age    = k %/% 32          # k = 0-based entry index
hh     = (k %/% 2) %% 16   # household type
window = k %% 2            # 0 -> Period 0..7, 1 -> Period 8..12
```

```{r}
# canivt exposes the geography count and member-order mapping internally;
# the codebook gives the human labels and the canonical DGUID per geography:
meta <- ivt_metadata(path)
head(data.frame(member_id = meta$geographies$member_id,
                name      = meta$geographies$name,
                dguid     = meta$geographies$dguid))
```

## A data page

Each directory entry points at a *page*:

```
[4-byte header][presence bitmap (256 bytes)][dense value array]
```

The first header byte encodes the value layout:

* low nibble `0x4` → `int32` cells, `0x2` → `int16` cells;
* high nibble `0x8` → values follow the presence bitmap directly, `0xa` → a
  `0xFF` separator run plus a 2-byte prefix precede the values.

So `84` = int32/plain, `a4` = int32/sep, `a2` = int16/sep, `82` = int16/plain.
Large geographies (big counts) use int32; small ones use int16.

## Presence bitmap

The 256-byte bitmap says which of the inner cells are stored. It is a
**positional, dimension-padded** map:

* 32-byte rows = one Period (row `r` → period `window*8 + r`);
* within a row, Statistics occupy 8-byte blocks at `stat*8` (the 4th stat slot
  is padding);
* within a block, Housing indicators 0..5 are used (6, 7 are padding) — **but the
  two bytes of each adjacent housing pair are stored byte-swapped**, so the
  presence byte for housing `h` is read at `stat*8 + bitwXor(h, 1)`;
* each presence byte's bits 7..1 flag Tenure members 0..6 (bit 0 is padding).
  `0xFE` therefore means "all 7 tenure values present", and `popcount(byte)` is
  the number of values stored for that group.

```{r, eval = TRUE}
# bits 7..1 of 0xFE flag the 7 tenure members; bit 0 is padding
present <- bitwAnd(bitwShiftR(0xFE, 7:1), 1L)
present
#> [1] 1 1 1 1 1 1 1
```

## Value codec

Values are stored **densely over the present cells** (zeros omitted), as
unaligned little-endian integers of the page's width, in plain
`(period, stat, housing, tenure)` order. Only the presence *bytes* are
pair-swapped — the value stream is not. To decode a page you walk the groups
period-outer → stat → housing-inner, and within each present group emit tenure
`t = 0..6` for the set bits (`t` uses bit `7 - t`), reading one integer per set
bit.

A useful generalisation (it also holds for the 1991 census IVTs): the presence
granularity is the innermost dimension, the "all present" marker is `2^n − 2`
over that bit width (Tenure `n = 7` → `0xFE`), and the bitmap is padded to fixed
per-dimension strides.

```{r}
# read_ivt() applies the codec to every page of every geography; the result is
# a long tibble of one value per row, keyed by 1-based member ids:
head(tab$cells)
```

## The codebook: labels, geographic ids, footnotes

The end of the file holds the codebook. For each dimension it stores several
parallel, **member-ordered** arrays of length-prefixed ("Pascal": one length
byte then that many text bytes, `0x00`-separated) **latin-1** strings:

* member ordinals, then the member **name** in English, then French;
* for Geography additionally: the level name, abbreviations, the classification
  code, and the full **DGUID** (`2021A000011124` = Canada, `2021A000210` =
  Newfoundland and Labrador, …) — the canonical geographic identifier, aligned
  with the geography index order.

Member names carry **leading-space indentation** that encodes the member
hierarchy (`Owner` → `  With mortgage`). `canivt` keeps the raw label and can
report the implied depth.

```{r}
# labelled, tidy long table (joins cells to the codebook)
ivt_tidy(tab)

# export the non-table metadata (members, geographies, footnotes, identity)
ivt_write_metadata(tab, "98100241_metadata")
```

Footnotes use a looser framing than the member arrays; `canivt` extracts them
best-effort by anchoring on the `Footnote N` (English) and `Renvoi N` (French)
markers:

```{r}
str(meta$footnotes[[1]])
```

## How `canivt` maps to the format

| Region | Function(s) | Source file |
|--------|-------------|-------------|
| Header identity & dimension names | `ivt_metadata()` | `R/codebook.R` |
| Geography index / page directory | (internal) `read_ivt()` | `R/container.R` |
| Page presence + value codec | (internal) `read_ivt()` | `R/decode.R` |
| Codebook labels, DGUIDs, footnotes | `ivt_metadata()` | `R/codebook.R` |
| Tidy / write outputs | `ivt_tidy()`, `ivt_write_*()` | `R/read.R`, `R/write.R` |

## Validation

The parser is validated against the StatCan CSV download for the reference table:
all 166 geographies, every DGUID present, dimension member counts
`166, 9, 16, 13, 3, 6, 7`, 7,489,464 decoded cells, and Canada's published tenure
totals `14687350, 9787420, 5870875, 3916550, 4899925, 576625, 4323300`.

## Other format variants

The 1991 census Beyond 20/20 files are the same container *family* with
different constants (page marker `82 01 80 08`, 1-byte note lengths, geography
codes inline in the label as `"(code)"`, presence keyed to Sex). They are **not**
yet supported; `read_ivt()` rejects unrecognised files. See
`system.file("notes/ivt-format.md", package = "canivt")` for the full spec,
including what differs for 1991.
