---
title: "The Beyond 20/20 IVT file format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Beyond 20/20 IVT file format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette documents the Statistics Canada *Beyond 20/20* `.ivt` binary
format as reverse-engineered for `canivt`, and shows how the package's functions
map onto each part of the file. It is aimed at anyone who wants to understand,
validate, or extend the parser. For day-to-day use see `?read_ivt`. The
authoritative, more detailed spec ships with the package:
`system.file("notes/ivt-format.md", package = "canivt")`.

## Three layouts, one container

Every `.ivt` starts with the signature `04 00 20 00` and is one contiguous
binary. `canivt` decodes two **container families** plus a legacy variant, all
validated cell-for-cell against the StatCan downloads:

* **Family 1** — *per-geography* page directories at a fixed stride. Reference
  table **98-10-0241** (housing indicators by tenure). 166 geographies.
* **Family 2** — a *single* contiguous page directory; 4 geographies per page.
  Reference table **98-10-0023** (age × gender, down to dissemination areas).
  63,404 geographies, 14.5 M cells.
* **Legacy (pre-2016)** — the 1991 census layout, e.g. **1003011** (E9101,
  population by single year of age × sex). Same container as family 2, but with
  `int16`/`int32` value pages and a pre-DGUID, bilingual, inline codebook.

`read_ivt()` auto-detects which one via `ivt_family()` and the header, so the
same call handles all three:

```{r}
library(canivt)
tab <- read_ivt(ivt_download("98100023"))      # family 2
tab
```

## File layout at a glance

The high-level region order is the same for every IVT: a small **header**, the
**page directory**, the **value pages** (which dominate — typically ~85–90 % of
the file), then the **codebook** (geography identifiers/labels and the data
dimension members), with **footnotes** near the tail.

```{r layout-plot, eval = TRUE, fig.width = 7, fig.height = 2.6, message = FALSE, fig.alt = "Tile plot of the IVT byte layout: a tiny header and page directory, a dominant value-pages region, then the geography codebook, footnotes and dimension blocks at the tail."}
library(ggplot2)

# Representative byte regions of the family-2 reference table 98-10-0023
# (142,016,485 bytes). Every IVT follows this order; the value pages dominate.
regions <- data.frame(
  region = c("Header", "Page directory", "Value pages (data)",
             "Geography codebook", "Footnotes", "Dimension blocks (Age/Gender)"),
  start  = c(        0,    35950,    167038, 124290000, 141893000, 141896000),
  end    = c(    35950,   162750, 124290000, 141893000, 141896000, 142016485)
)
regions$region <- factor(regions$region, levels = regions$region)
regions$pct <- round(100 * (regions$end - regions$start) / max(regions$end), 1)

ggplot(regions, aes(xmin = start / 1e6, xmax = end / 1e6,
                    ymin = 0, ymax = 1, fill = region)) +
  geom_rect(colour = "white", linewidth = 0.4) +
  scale_x_continuous("file offset (MB)", expand = c(0, 0)) +
  scale_y_continuous(NULL, breaks = NULL, expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set2", name = NULL,
                    labels = paste0(levels(regions$region),
                                    "  (", regions$pct, "%)")) +
  labs(title = "IVT file layout — table 98-10-0023 (142 MB)",
       subtitle = "Value pages are the bulk; all metadata sits at the head and tail") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right", panel.grid = element_blank(),
        plot.title.position = "plot")
```

So roughly 87 % of the file is data; the geography codebook is the second-largest
region (~12 %), and the header, directory, footnotes and per-dimension member
blocks are slivers. All multi-byte integers are **little-endian**, and value runs
frequently begin on **odd (unaligned) byte offsets** — a property that defeats
naive aligned scans. Byte offsets in the parser are 0-based to match the layout.

## Reading the layout from the header

The fixed header points at every major section, so the layout can be read up
front without scanning for markers. `ivt_f2_header_layout()` (family 2 / legacy)
returns these as byte offsets:

| header offset | section |
|---|---|
| `u32 @32` | dimension descriptor (per-dimension count, type, name + title) |
| `u32 @40` / `u32 @48` | French / English title blocks — **set in the legacy format, zero in the modern one** (the version indicator) |
| `u32 @552` | geography field/attribute count (11 modern / 12 legacy) |
| `u16 @558` | page-directory start |
| `u32 @572` | codebook region start |

The page directory is a contiguous run of `[u32 offset][u16 size][u16 size]`
records (the two sizes agree), in geography member order. Each record points at a
*page* of value data.

## Dimensions

The dimension descriptor declares each dimension uniformly. For 98-10-0023:
Geography (63,404) × Age (128) × Gender (3). The first dimension always indexes
the pages; the rest are the "inner" dimensions stored within each page, with the
**last** dimension (here Gender) as the unit of value storage.

```{r}
meta <- ivt_metadata(ivt_download("98100023"))
data.frame(dimension = meta$dimension_names,
           members   = meta$dimension_counts)
```

## Family 2: a single page directory (98-10-0023)

Each directory record is a page that packs **4 geographies**:

```
[4-byte marker][4 × 64-byte presence records][0xFF trailer][dense value run]
```

* the **marker**'s low nibble is the value width: `0x8` → float64, `0x4` →
  int32, `0x2` → int16 (so age counts are int16 but the average/median-age
  statistics are float64);
* each geography's **presence** is a 64-byte record stored **byte-pair-swapped**
  — swap adjacent bytes, then read a positional nibble per member, with the three
  genders Total/Men/Women at bits 3/2/1 (the all-present marker is `0xE`);
* **values** are dense little-endian over the present `(member, gender)` cells,
  member-major / gender-inner. Only non-zero cells are stored — the StatCan CSV
  publishes the zeros, so a missing cell means 0.

`ivt_f2_geo_attributes()` decodes the full per-geography codebook: name, DGUID,
geographic level and type (+ abbreviation), two geocodes, the data-quality flag
and note, and the non-response rate — all exact for 63,404 geographies.

## Family 1: per-geography directories (98-10-0241)

The original layout instead gives **each geography its own page directory** at a
fixed stride:

```
[0]            header: identity + dimension names
[37167]        geography index: one page-directory per geography, stride 0x1000
~1.08M..~55.4M value pages
~56.93M..EOF   footnote legend + codebook
```

> **Gotcha.** The directories are grouped 8 to a `0x8000` region (288 entries
> used per `0x1000` slot, the rest zero-padded). Striding by `0x8000` instead of
> `0x1000` silently reads only every 8th geography — 21 of 166. The correct
> stride is `0x1000`.

Here the page presence is a **256-byte positional, dimension-padded** bitmap:
32-byte rows per Period; Statistics in 8-byte blocks; Housing indicators 0..5
within a block (the two bytes of each adjacent housing pair are **byte-swapped**,
so housing `h` is read at `stat*8 + bitwXor(h, 1)`); and each presence byte's
bits 7..1 flag the 7 Tenure members (bit 0 padding). `0xFE` therefore means "all
7 tenure values present".

```{r, eval = TRUE}
# bits 7..1 of 0xFE flag the 7 tenure members; bit 0 is padding
bitwAnd(bitwShiftR(0xFE, 7:1), 1L)
#> [1] 1 1 1 1 1 1 1
```

A useful generalisation across families: the presence granularity is the
**innermost** dimension, the "all present" marker is `2^n − 2` over that bit
width, the per-dimension strides are fixed, and one byte of each adjacent pair is
swapped.

## The codebook: labels, geographic ids, footnotes

The tail of the file holds the codebook: for each dimension, several parallel,
**member-ordered** arrays of length-prefixed ("Pascal": one length byte then that
many text bytes) strings — member ordinals, the name (English then French) and,
for Geography, the level, abbreviations, classification code, and full **DGUID**
(`2021A000011124` = Canada). Labels are encoded in **Windows-1252**: the
`0x80–0x9F` block carries real punctuation (e.g. `0x92` = the curly apostrophe in
`Tla'amin Lands`), so the byte-class test that frames Pascal records must accept
it — otherwise such a label aborts a record and splits the array mid-stream.

```{r}
# labelled long table; read_ivt(geo_attributes = TRUE) attaches the full
# geography attribute table so tidy can label by name + level.
tab <- read_ivt(ivt_download("98100023"), geo_attributes = TRUE)
ivt_tidy(tab)
```

Footnotes sit near the codebook. In the modern format each is framed and tagged
by a leading `Footnote N` / `Renvoi N` marker; `canivt` recovers each as a
maximal run of text bytes beginning with that marker. The legacy format instead
stores them as `(N) text` lines under a `Footnotes` section header in one notes
block.

```{r}
str(meta$footnotes[[1]])
length(meta$footnotes)
```

## Legacy 1991 variant (1003011)

The 1991 census files are the same family-2 container with pre-DGUID conventions,
all handled by `read_ivt()`:

* value pages are `int16`/`int32` (markers `0x82`/`0x84`); cell decode is exact
  for the scraped ground-truth geographies;
* geography is a single inline block per chunk of `"<name> (<GEOUID>) <flag>"` —
  a **bilingual** name (`Newfoundland | Terre-Neuve`), a bare **GEOUID** (a
  shortened DGUID without the year/area-type prefix), and a data-quality flag;
* the table identity is stored *out of line* (the `@40`/`@48` header pointers),
  and footnotes are the `(N) text` form described above.

```{r}
leg <- read_ivt(ivt_download("1003011"))   # 41,859 geographies, Age(110) × Sex(3)
ivt_tidy(leg)
```

## How `canivt` maps to the format

| Region | Function(s) | Source file(s) |
|--------|-------------|----------------|
| Header identity, dimension descriptor, layout | `ivt_metadata()`, `ivt_f2_header_layout()` | `R/codebook*.R` |
| Page directory / geography index | (internal) `read_ivt()` | `R/container*.R` |
| Page presence + value codec | (internal) `read_ivt()` | `R/decode*.R` |
| Codebook labels, DGUIDs/GEOUIDs, geography attributes, footnotes | `ivt_metadata()`, `ivt_f2_geo_attributes()` | `R/codebook*.R`, `R/read-f2.R` |
| Tidy / write outputs | `ivt_tidy()`, `ivt_write_*()` | `R/read.R`, `R/write.R` |

## Validation

`canivt` is validated against the StatCan CSV/metadata downloads: family 1
(98-10-0241) all 166 geographies and 7,489,464 cells exact; family 2 (98-10-0023)
all 63,404 geographies and 14.5 M cells exact, plus every geography attribute and
the Age/Gender labels; and the 1991 table (1003011) cell-exact for all scraped
ground-truth geographies. See the `tests/` directory and the notes file for the
exact figures.
